---
title: "Assignment 3: Data Exploration"
author: "Elise Harrigan"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. 
```{r,  message = FALSE}

library(tidyverse)

Neonics <- read.csv("~/Desktop/Duke/Spring 2021/EnvDataAnalytics_872/Environmental_Data_Analytics_2021/Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactor=TRUE)

Litter <- read.csv("~/Desktop/Duke/Spring 2021/EnvDataAnalytics_872/Environmental_Data_Analytics_2021/Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)

```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer:Neonicotinoids are used to keep crops safe and healthy from insects that may interfere with the productivity or viability of the crop. It is an insecticide that directly targets certain species of insects that tend to feed or growth in field crops. With the use of insecticides, there can be unforeseen consequences as pollinating insects, which are necessary, might also be negatively affected by the usage of the product. 

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Forest litter and woody debris is critical to determine forest health. In the case of Colorado, a dry and arid climate,  too much forest litter and debris can lead to increase in wildfires. If there is a lighting storm or manmade fires started, a build up of litter can dry out and create mass fire outbreaks.

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
*Sampling of woody debris and litter occurs in sites over 2m tall in woody vegetation
*Sites with forested tower airsheds the litter sampling plots are 20 40x40m tower plots and 26 20x20m plots.There is one elevated trap and one ground trap every 400 m^2 plot area which results in 1-4 trap pairs per plot. 
* In plots with greater than 50% of forest cover, traps are randomized and utilized on a grid cell location. For plots less than 50% coverage or patchy vegetation, traps are placed in targeted areas based on the vegetaion. 


## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}


dim(Neonics)
#Neonics has 4623 rows and 30 columns


```

6. Using the `summary` function on the "Effects" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
summary(Neonics$Effect)


```

> Answer: the most common effects that are studied are Population, Mortality and Feeding behavior. These are most common because it is important to know how many insects you are dealing iwth, the mortality rate and how they are able to feed and continue to grow in their life cycle.   

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(Neonics$Species.Common.Name)

```

> Answer: The most common thread among the top species are they they are pollinators. Insecticides are dangerous to pollinator communities such as honey bee, bumble bee and wasp. Although insecticides are designed to rid the crops of infestation of pests, it has a negative effect on other important insects such as bees. 
                         Honey Bee                     Parasitic Wasp 
                               667                                285 
             Buff Tailed Bumblebee                Carniolan Honey Bee 
                               183                                152 
                        Bumble Bee                   Italian Honeybee 
                               140                                113 
                               
8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.)
head(Neonics$Conc.1..Author.)

```

> Answer: The class is a character. It is not numeric because it contains more than just numbers. 



## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}
library(ggplot2)
ggplot(Neonics, aes(Publication.Year)) + 
  geom_freqpoly()
summary(Neonics$Publication.Year)

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
ggplot(Neonics, aes(Publication.Year, colour=Test.Location)) + 
  geom_freqpoly()


```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test location is in the lab location, peaking in 2015. Historically natural field sites were also common, especially from the mid 1990s to 2010 where there was a major spike in field locations. Both lab and field sites have peaks and dips over time.There are very little publications using artifical sites and none using undeterminable sites. But recently the publications have dropped in all test locations. One reason for this could be the data hasn't been provided for 2019-2020 but also COVID-19 could have an effect on the amount of testing possible. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}
summary(Neonics$Endpoint)
 
counts <- table(Neonics$Endpoint)

barplot(counts, main="Endpoint Counts",
        xlab="Endpoints")

```

> Answer: 
>The most common endpoints are LOEL and NOEL. LOEL is terrestrial and stands for Lowest-observable-effect-level: lowest dose that produced significantly different values. NOEL is also terrestrial and stands for No-observable-effect-level: highest dose producing effects not significantly different from responses.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
class(Litter$collectDate)
head(Litter$collectDate) #Yes it is a date.

library("lubridate")

unique(Litter$collectDate)

```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$plotID)
 

summary(Litter$plotID)



```

> Answer: There were 12 plots sampled at Niwot Ridge. The unique function gives you the value of the number of plots taken and summary gives you summary statistics on the variable. In this case, summary gave how many of each type of plot there were and unique gives the "unique value" without repeating. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}

func_group_counts <- table(Litter$functionalGroup)

barplot(func_group_counts, main="Type of Litter at Niwot Ridge",
        xlab="Type", ylab="Count")

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r, fig.height = 3, fig.width = 4}

ggplot(Litter) +
  geom_boxplot(aes(x=dryMass, y=functionalGroup))

ggplot(Litter) +
  geom_violin(aes(x = dryMass, y = functionalGroup))

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The boxplot shows more details on the range of variables. You can see the outliers and where the mean of the data fall. The violin plot does not accurately describe the fairly equal distribution across the group types. 

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Needles has the highest biomass. 